### 2018-03-17 paper/4-build.tex
S4 and S7 tweaks

As per the definition of minimality~\ref{def-minimal}, a minimal build
system must \underline{rebuild only out-of-date keys} and at most once. The only
way to achieve the ``at most once'' requirement while producing a correct build
result (\S\ref{sec-build-correctness}) is to \underline{build all keys in an
order that respects their dependencies}.

\vspace{1mm}
We have underlined two different phrases above, and tackle each aspect separately.

@@

The topological approach pre-computes a linear order, which when followed, ensures
the \hs{build} is correct regardless of the initial \hs{store}. Given a function
from a key to its dependencies, and the output \hs{key}, you can compute the
linear order by first taking the transitive closure of \hs{key}'s dependencies,
and then computing its topological sort. However, as we have seen
in~\S\ref{sec-deps}, we can only extract dependencies from an applicative task,
which requires the build system to choose \hs{c}~\hs{=}~\hs{Applicative}, ruling
out dynamic dependencies.

@@

The topological approach has two downsides: it is limited to \hs{Applicative}
build systems and requires a fresh topological sort each time.  So, while the
actions themselves may be incremental (i.e. unnecessary tasks will not be performed),
the pre-processing is not. We may try to incrementalise the topological sort by
storing the topological order between build runs and assume it to
be correct, but if the build determines it is wrong, fix it up.

@@

An alternative approach, utilised by the \hs{busy} build system
(\S\ref{sec-general-build}), is to simply build dependencies when they are
requested. By combining that with a transient log of which keys have already
been built, you can obtain a minimal build system.

This approach requires that a task may be started, then during that execution
another task will have to be run. Assuming a an IO-driven task structure,
that requires suspend a running task, which can be done with cheap
green threads and blocking (the original approach of \Shake) or with a
continuation-passing style (what \Shake does now). An alternative approach to
suspending a task is to abort it and restart it again later, at the cost
of doing additional work.

@@

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell dirty if the user modifies it.
When rebuilding, if a cell only depends on clean cells it is skipped, otherwise
it is rebuilt and is marked dirty so that the cells that depend on it are
subsequently rebuilt too.

@@

\Make uses file modification times, and compares files to their
dependencies, which can be thought of as a dirty bit which is set when
a file is newer than its dependencies. The interesting property of
this dirty bit is that it is not under control of \Make; rather it is
existing file-system information that is repurposed. In particular,
rebuilding a file automatically clears its dirty bit, and
automatically sets the dirty bit of the nodes depending on it. One
thing \Make does require is that file timestamps only go forward in
time -- something that can be violated by backup software.

@@

With a dirty bit it is simple to achieve minimality. However, to achieve early cutoff
(\S\ref{sec-background-shake}) it would be important to not set the dirty bit
after a computation that did not change the value. \Excel could use this
approach, but does not. In contrast, \Make cannot implement early cutoff nicely -- to
do so it would have to mark the node clean (so it would not rebuild in the next
run) and at the same time not mark the things it depends on dirty -- an
impossible task with only the ability to update to the latest modification time.
It can approximate early cutoff by not touching the result file, and not marking it clean,
but then it reruns in every subsequent build.

>>>>

As per the definition of minimality~\ref{def-minimal}, a minimal build
system must \textbf{rebuild only out-of-date keys} and at most once. The only
way to achieve the ``at most once'' requirement while producing a correct build
result (\S\ref{sec-build-correctness}) is to \textbf{build all keys in an
order that respects their dependencies}.

\vspace{1mm}
We have bolded two different phrases above, and tackle each aspect separately.

@@

The topological approach pre-computes a linear order, which when followed, ensures
the \hs{build} is correct regardless of the initial \hs{store}. Given a function
from a key to its dependencies, and the output \hs{key}, you can compute the
linear order by first finding the reachable dependencies of \hs{key},
and then computing a topological sort. However, as we have seen
in~\S\ref{sec-deps}, we can only extract dependencies from an applicative task,
which requires the build system to choose \hs{c}~\hs{=}~\hs{Applicative}, ruling
out dynamic dependencies.

@@

The topological approach has two downsides: it is limited to \hs{Applicative}
build systems and requires a fresh topological sort each time.  So, while the
actions themselves may be incremental (i.e. unnecessary tasks will not be performed),
the pre-processing is not. We can try to incrementalise the topological sort by
storing the topological order between build runs and assume it to
be correct, but if the build discovers it is wrong, fix it up.

@@

An alternative approach, utilised by the \hs{busy} build system
(\S\ref{sec-general-build}), is to simply build dependencies when they are
requested. By combining that with a transient set of which keys have already
been built, you can obtain a minimal build system.

This approach requires that a task may be started, then during that execution
another task will have to be run. Assuming a an IO-driven task structure,
that requires suspending a running task, which can be done with cheap
green threads and blocking (the original approach of \Shake) or using
continuation-passing style (what \Shake does currently). An alternative approach to
suspending a task is to abort it and restart it again later, at the cost
of doing additional work.

@@

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell dirty if the user modifies it.
When rebuilding, if a cell only depends on clean cells it is skipped, otherwise
it is rebuilt and marked dirty so that the cells that depend on it are
subsequently rebuilt too.

@@

\Make uses file modification times, and compares files to their
dependencies, which can be thought of as a dirty bit which is set when
a file is newer than its dependencies. The interesting property of
this dirty bit is that it is not under the control of \Make; rather it is
existing file-system information that has been repurposed. In particular,
modifying a file automatically clears its dirty bit, and
automatically sets the dirty bit of the nodes depending on it. One
thing \Make does require is that file timestamps only go forward in
time -- something that can be violated by backup software.

@@

With a dirty bit it is simple to achieve minimality. However, to achieve early cutoff
(\S\ref{sec-background-shake}) it would be important to not set the dirty bit
after a computation that did not change the value. \Excel could use this
approach, but does not. In contrast, \Make cannot implement early cutoff nicely -- to
do so it would have to mark the node clean (so it would not rebuild in the next
run) and at the same time not mark the things it depends on dirty -- an
impossible task with only the ability to update to the latest modification time.
\Make can approximate early cutoff by not modifying the result file, and not marking it clean,
but then it will rerun in every subsequent build.

### 2018-03-17 paper/7-related.tex
S4 and S7 tweaks

While not typically considered build systems, self-adjusting computation is a
well studied area, and in particular the contrast between different formulations
has been well studied~\cite{acar2007selfadjusting}.

Self-adjusting computations can automatically adjust to an external change
to their inputs. A classic example is a self-adjusting sorting algorithm, which
can efficiently (in $O(\log{n})$ time where $n$ is the length of the input)
recalculate the result given an incremental change of the input. While very
close to build systems in spirit, self-adjusting computations are mostly used
for in-memory computation and rely on the ability to dynamically allocate new
keys in the store for sharing intermediate computations~--~an intriguing feature
rarely seen in build systems (\Shake's oracles~\S\ref{sec-polymorphism} can be
used to model this to a limited degree).

A lot of research has been dedicated to finding efficient data structures and
algorithms for self-adjusting computations~--~we plan to explore this
untapped (by build systems) resource in~future.

>>>>

While not typically considered build systems, self-adjusting computation is a
well studied area, and in particular the contrast between different formulations
has been thoroughly investigated~\cite{acar2007selfadjusting}.

Self-adjusting computations can automatically adjust to an external change
to their inputs. A classic example is a self-adjusting sorting algorithm, which
can efficiently (in $O(\log{n})$ time where $n$ is the length of the input)
recalculate the result given an incremental change of the input. While very
close to build systems in spirit, self-adjusting computations are mostly used
for in-memory computation and rely on the ability to dynamically allocate new
keys in the store for sharing intermediate computations~--~an intriguing feature
rarely seen in build systems (\Shake's oracles~\S\ref{sec-polymorphism} can be
used to model this feature to a limited degree).

A lot of research has been dedicated to finding efficient data structures and
algorithms for self-adjusting computations~--~we plan to investigate how these
insights can be utilised by build systems as future work.
